Looking for a Signal in the Noise: Revisiting Obesity and the Microbiome

### 2016-06-03 submission/Sze_Obesity_mBio_2016.Rmd
Editing manuscript

Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity. This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets. The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis. Alpha diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances. Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%. A power analysis demonstrated that `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity. When models trained on one dataset were then tested using the other 9 datasets, the median accuracy varied between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`% (median=`r 100*median(test_summary$accuracy)`%). Although there is statistical support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.

@@

There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans. The first such study used genetically obese mice and observed the ratio of the relative abundances of Bacteroidetes to Firmicutes (B:F) was lower in obese mice than lean mice [@ley_obesity_2005]. Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009]. They also showed that the relative abundance of Bacteroidetes and Firmicutes increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006]. Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since direct manipulation of the communities yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016]. The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

@@

***Meta-analysis***

[@ross_16s_2015; @zupancic_analysis_2012; @goodrich_human_2014; @wu_linking_2011; @escobar_gut_2014; @baxter_microbiota-based_2016; @schubert_microbiome_2014; @zeevi_personalized_2015; @human_microbiome_project_consortium_structure_2012; @turnbaugh_obesity-associated_2006]
```{r}
alpha_sig <- alpha[alpha$p.value < 0.05,]
sig_counts <- table(alpha_sig$metric)

@@

***Alpha diversity analysis.***
We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample. Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the `r number_string[n]` studies. The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study. We identified `r sum(alpha$p.value < 0.05)` P-values less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1). These results largely match those of the Walters and Finucane re-analysis studies. Interestingly, although only `r number_string[sig_counts["shannon"]]` of the 10 studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant. We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`). Although there was a significant relationship between these metrics and diversity and obesity status, the effect size was quite small. The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity. There were no significant associations when we pooled the phylum-level metrics across studies. These results indicate that obese individuals do have a statistically significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant.

@@

***Relative risk.***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on whether an individual's alpha diversity metrics were below or above the median metric for that study. The results using relative risk largely matched those of using the untransformed alpha diversity data. Across the number_string[n] studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant difference in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

@@

***Beta diversity analysis.***
Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study. We then used AMOVA to test for a significant differences between the structure of non-obese and obese individuals. The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05). Because it was not possible to ascertain the directionality of the difference in community structure nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.

@@

***Development of a microbiome-based classifier of obesity.***
The Walters re-analysis study suggested that it was possible to classify individuals as being obese or non-obese based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The the observed AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this to identify a threshold to calculate the accuracy of the models when applied to the other number_string[n-1] datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

@@

***Power and Sample Size Estimate Simulations.***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S3-S8). Although there is no biological rationale for these effect sizes, they represent a range that is plausible. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, and 15% difference in Shannon index, the median sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)`  individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, and  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

@@

Our meta-analysis helps to provide a clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity. We performed an extensive literature review of the existing studies on the microbiome and obesity and perform a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics. We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome is limited. Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes. Considering these datasets are among the largest published, it appears that most microbiome studies are underpowered to detect differences in alpha diversity.

@@

Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference. Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from 3 studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from 5 studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012]. Our analysis included 4 of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded 3 of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012]. The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies. Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study. To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available. Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g. subject's sex or age).

Similar to our study, the Walters analysis [@walters_meta-analyses_2014], the authors generated Random Forest machine learning models to differentiate between non-obese and obese individuals. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.

Although we failed to find an effect it is not realistic to necessarily state that there is no microbiome impact on obesity. There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005]. These studies appear to conflict with the observations using human subjects. Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity. Alternatively, it could be that the involvement of the microbiome in obesity is at the level of a common set of metabolites that can be produced from different structures of the microbiome.

@@

***Literature Review and Study Inclusion.***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplement and in figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_core_2009; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The number_string[n] studies are summarized in Tables 1 and 2.

@@

**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups Based on Evenness (A), Richness (B), or the Relative Abundance of Bacteroidetes (C) and Firmictues (D).**

**Figure S2: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of Bacteroidetes (C) and Firmictues (D).**

@@

**Figure S6: Power (A) and sample size simulations (B) for the relative abundance of Bacteroidetes for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of Firmicutes for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

>>>>

Two recent studies have re-analyzed published data and found that when datasets are analyzed independently there was limited support the widely accepted hypothesis that changes in the microbiome are associated with obesity. This hypothesis was reconsidered by increasing the number of data sets and pooling the results across the individual datasets. The Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines were applied to identify 10 studies for an updated and more synthetic analysis. Alpha diversity metrics and the relative risk of obesity based on those metrics were used to identify a limited number of significant associations with obesity; however, when the results of the studies were pooled using a random effects model significant associations were observed between Shannon diversity, number of observed OTUs, and Shannon evenness and obesity status. They were not observed for the ratio of *Bacteroidetes* and *Firmicutes* or their individual relative abundances. Although these tests yielded small P-values, the difference between the Shannon diversity index of non-obese and obese individuals was `r alpha_effect_size["shannon"]`%. A power analysis demonstrated that only `r number_string[well_powered]` of the studies had sufficient power to detect a 5% difference in diversity. When Random Forest machine learning models were trained on one dataset and then tested using the other 9 datasets, the median accuracy varied between `r 100*range(test_summary$accuracy)[1]` and `r 100*range(test_summary$accuracy)[2]`% (median=`r 100*median(test_summary$accuracy)`%). Although there was support for a relationship between the microbial communities found in human feces and obesity status, this association is relatively weak and its detection is confounded by large interpersonal variation and insufficient sample sizes.

@@

There have been several studies that report observing a link between the composition of microbiome and obesity in animal models and in humans. The first such study used genetically obese mice and observed the ratio of the relative abundances of *Bacteroidetes* to *Firmicutes* (B:F) was lower in obese mice than lean mice [@ley_obesity_2005]. Translation of this result to humans by the same researchers did not observe this effect, but did find that obese individuals had a lower diversity than lean individuals [@turnbaugh_core_2009]. They also showed that the relative abundance of *Bacteroidetes* and *Firmicutes* increased and decreased, respectively, as obese individuals lost weight while on a fat or carbohydrate restricted diet [@ley_microbial_2006]. Two re-analysis studies interrogated previously published microbiome and obesity data and concluded that the previously reported differences in community diversity and B:F among non-obese and obese individuals could not be generalized [@walters_meta-analyses_2014; @finucane_taxonomic_2014]. Regardless of the results using human populations, mechanistic studies using animal models that were manipulated with antibiotics or colonization with varied communities appears to support the association since these manipulation yielded variation in animal weight [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016]. The purported association between the differences in the microbiome and obesity have been widely repeated with little attention given to the lack of a clear signal in human cohort studies.

@@

***Literature Review and Study Inclusion.***
We followed the Preferred Reporting Items for Systematic Reviews and Meta-Analyses (PRISMA) guidelines to identify studies to include in our meta-analysis [@moher_preferred_2010]. A detailed description of our selection process and the exact search terms are provided in the Supplemental Text and in Figure 1. Briefly, we searched PubMed for original research studies that involved studying obesity and the human microbiome. The initial search yielded 187 studies. We identified number_string[n] additional studies that were not designed to explicitly test for an association between the microbiome and obesity. We then manually curated the 196 studies to select those studies that included BMI and sequence data. This yielded 10 eligible studies. An additional study was removed from our analysis because no individuals in the study had a BMI over 30. Among the final 10 studies, 3 were from identified from our PubMed search [@turnbaugh_obesity-associated_2006; @zupancic_analysis_2012; @escobar_gut_2014], 5 were originally identified from the 9 studies that did not explicitly investigate obesity but included BMI data [@zeevi_personalized_2015; @ross_16s_2015; @goodrich_human_2014; @wu_linking_2011; @human_microbiome_project_consortium_structure_2012], and two datasets were used [@schubert_microbiome_2014; @baxter_microbiota-based_2016] because these publications did not specifically look for any metabolic or obesity conditions but had control populations and enabled us to help mitigate against publication biases associated with the bacterial microbiome and obesity. The `r number_string[n]` studies are summarized in Table 1.

```{r}
alpha_sig <- alpha[alpha$p.value < 0.05,]
sig_counts <- table(alpha_sig$metric)

@@

***Alpha diversity analysis.***
We calculated the Shannon diversity index, richness, and Shannon evenness,  the relative abundance of *Bacteroidetes* and *Firmicutes*, and the ratio of their relative abundance (B:F) for each sample. Once we transformed each of the six alpha diversity metrics to make them normally distributed, we used a t-test to identify significant associations between the alpha diversity metric and whether an individual was obese for each of the `r number_string[n]` studies. The B:F and the relative abundance of *Firmicutes* were not significantly associated with obesity in any study. We identified `r number_string[sum(alpha$p.value < 0.05)]` P-values that were less than 0.05: `r number_string[sig_counts["sobs"]]` studies indicated obese individuals had a lower richness, `r number_string[sig_counts["shannon"]]` studies indicated a significantly lower diversity, `r number_string[sig_counts["shannoneven"]]` study indicated a significantly lower evenness, and `r number_string[sig_counts["bacteroidetes"]]` study indicated a significantly higher relative abundance of *Bacteroidetes* (Figures 2 and S1). These results largely match those of the Walters and Finucane re-analysis studies. Interestingly, although only `r number_string[sig_counts["shannon"]]` of the `r number_string[n]` studies observed the previously reported association between lower diversity and obesity, the other studies appeared to have the same trend, albeit the differences were not statistically significant. We used a random effects linear model to combine the studies using the study as the random effect and found statistical support for decreased richness, evenness, and diversity among obese individuals (all P<`r max(alpha_composite[alpha_composite$p <0.05, "p"])`). Although there was a significant relationship between these metrics and obesity status, the effect size was quite small. The obese individuals averaged `r alpha_effect_size["sobs"]`% lower richness, `r alpha_effect_size["shannoneven"]`% lower evenness, and `r alpha_effect_size["shannon"]`% lower diversity. There were no significant associations when we pooled the phylum-level metrics across studies. These results indicate that obese individuals do have a statistically significant lower diversity than non-obese individuals; however, it is questionable whether the difference is biologically significant.

@@

***Relative risk.***
Building upon the alpha diversity analysis we calculated the relative risk of being obese based on an individual's alpha diversity metrics relative to the median metric for that study. The results using relative risk largely matched those of using the untransformed alpha diversity data. Across the `r number_string[n]` studies and six metrics, the only significant relative risk values were the richness, evenness, and diversity values from the Goodrich study (Figures 3 and S2). Again, although the relative risk values were not significant for other studies, the values tended to be above one. When we pooled the data using a random effects model, the relative risk associated with having a richness, evenness, or diversity below the median for the population was significantly associated with obesity (all P<`r max(relative_composite[relative_composite$p_value <0.05, "p_value"])`). The relative risks associated with alpha diversity were small. The relative risk of having a low richness was `r concatenate_rr(relative_composite["sobs",])`, low evenness was `r concatenate_rr(relative_composite["shannoneven",])`, and low diversity was `r concatenate_rr(relative_composite["shannon",])`. There were no significant difference in the phylum-level metrics. Again, the relative risk results indicate that individuals with a lower richness, evenness, or diversity are at statistically significant increased risk of being obese, it is questionable whether that risk is biologically or clinically relevant.

@@

***Beta diversity analysis.***
Following the approach used by the Walters and Finucane re-analysis studies, for each dataset we calculated a Bray-Curtis distance matrix to measure the difference in the membership and structure of the individuals from each study. We then used AMOVA to test for a significant differences between the structure of non-obese and obese individuals (Table 1). The Escobar, Goodrich, and Turnbaugh datasets indicated a significant difference in community structure (all P<0.05). Because it was not possible to ascertain the directionality of the difference in community structure nor perform a pooled analysis using studies that had non-overlapping 16S rRNA gene sequence regions it is unclear whether these differences reflect a broader, but perhaps small, shift in community structure between non-obese and obese individuals.

@@

***Development of a microbiome-based classifier of obesity.***
The Walters re-analysis study suggested that it was possible to classify individuals as being non-obese or obese  based on the composition of their microbiota. We repeated this analysis with additional datasets using OTU and genus-level phylotype data. For each study we developed a Random Forest machine learning model to classify individuals. Using ten-fold cross validation, the observed AUC values varied between `r min(otu$auc_cv)` and `r max(otu$auc_cv)` indicating a relatively poor ability to classify individuals (Figure 4A). So that we could test models on other datasets, we trained models using genus-level phylotype data for each dataset. The cross-validated AUC values for the models applied to the training datasets varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)`, again indicating a relatively poor ability to classify individuals from the original dataset (Figure 4B). For each model we identified the probability where the sum of the sensitivity and specificity was the highest. We then used this probability to define a threshold for calculating the accuracy of the models when applied to the other `r number_string[n-1]` datasets (Figure 5). Although there considerable variation in accuracy values for each model, the median accuracy for each model varied between `r min(test_summary$accuracy)` (Turnbaugh) and `r max(test_summary$accuracy)` (HMP) (median=`r median(test_summary$accuracy)`). When we considered the number of samples, balance of non-obese and obese individuals, and region within the 16S rRNA gene it was not possible to identify factors that predictably affected model performance. The ability to predict obesity status using the relative abundance of OTUs and genera in the communities is only marginally better than random. These results suggest that given the large diversity of microbiome compositions it is difficult to identify a taxonomic signal that can be associated with obesity.

@@

***Power and Sample Size Estimate Simulations.***
The inability to detect a difference between non-obese and obese individuals could be due to the lack of a true effect or because the study had insufficient statistical power to detect a difference because of insufficient sampling, large interpersonal variation, and unbalanced sampling of non-obese and obese individuals. To assess this, we calculated the power to detect differences of 1, 5, 10, and 15% in each of the alpha diversity metrics using the sample sizes used in each of the studies (Figures 6, S3-S8). Although there is no biological rationale for these effect sizes, they represent a range that is plausible. Only the Goodrich study  had power greater than 0.80 to detect a 5% difference in Shannon diversity and `r number_string[length(which_div_at_10)]` of the studies had enough power to detect a 10% difference (Figure 6). None of the studies had sufficient power to detect a 15% difference between B:F values (Figure S5). In fact, the maximum power among any of the studies to detect a 15% difference in B:F values was `r max(power[power$metric=="bf_ratio" & power$effect_size==0.15,"power"])`. Among the tests for relative risk, none of the studies had sufficient power to detect a Cohen's d of 0.10 and only `r number_string[sum(rr_power[rr_power$metric=="shannon" & power$effect_size==0.15,"power"] > 0.80)]` studies had sufficient power to detect a Cohen's d of 0.15. We next estimated how many individuals would need to have been sampled to have sufficient power to detect the four effect sizes assuming the observed interpersonal variation from each study and balanced sampling between the two groups. To detect a 1, 5, 10, or 15% difference in Shannon index, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="shannon" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="shannon" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="shannon" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in B:F values, the median required sampling effort per group was approximately `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(power[power$metric=="bf_ratio" & power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. To detect a 1, 5, 10, and 15% difference in relative risk values using Shannon diversity, the median required sampling effort per group was approximately `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.01,"balanced_n"]), 2)`, `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.05,"balanced_n"]), 2)`,  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.10,"balanced_n"]), 2)`, or  `r signif(median(rr_power[rr_power$metric=="bf_ratio" & rr_power$effect_size==0.15,"balanced_n"]), 2)` individuals, respectively. These estimates indicate that most microbiome studies are underpowered to detect modest effect sizes using either metric. In the case of obesity, the studies were underpowered to detect the `signif(range_effect_size, 1)[1]` to `r signif(range_effect_size, 1)[2]`% difference in diversity that was observed across the studies.

@@

Our meta-analysis helps to provide a clarity to the ongoing debate of whether or not there are specific microbiome-based markers that can be associated with obesity. We performed an extensive literature review of the existing studies on the microbiome and obesity and performed a meta-analysis on the studies that remained based on our inclusion and exclusion criteria. By statistically pooling the data from ten studies, we observed significant, but small, relationships between richness, evenness, and diversity and obesity status as well as the relative risk of being obese based on these metrics. We also generated Random Forest machine learning models trained on each dataset and tested on the remaining datasets. This analysis demonstrated that the ability to reliably classify individuals as being obese based on the composition of their microbiome was limited. Finally, we assessed the ability of each study to detect defined differences in alpha diversity and observed that most studies were underpowered to detect modest effect sizes. Considering these datasets are among the largest published, it appears that most microbiome studies are underpowered to detect differences in alpha diversity.

@@

Two previous reviews [@walters_meta-analyses_2014; @finucane_taxonomic_2014] have stated that there was not a consistent association between alpha diversity and obesity; however, neither of these studies made an attempt to pool the existing data together to try and harness the additional power that this would give and they did not assess whether the studies were sufficiently powered to detect a difference. Our analysis also used 16S rRNA gene sequence data from ten studies whereas the Finucane study used 16S rRNA gene sequence data from three studies [@turnbaugh_obesity-associated_2006; @ley_microbial_2006; @human_microbiome_project_consortium_structure_2012] and a metagenomic study [@Qin2010] and the Walters study used 16S rRNA gene sequence data from five studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011; @yatsunenko_human_2012]; two studies were included in both analyses [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012]. Our analysis included four of these studies [@turnbaugh_obesity-associated_2006; @human_microbiome_project_consortium_structure_2012; @zupancic_analysis_2012; @wu_linking_2011] and excluded three of the studies because they were too small [@ley_microbial_2006], only utilized metagenomic data [@Qin2010], or used short single read Illumina HiSeq data that has a high error rate making it untractable for *de novo* OTU clustering [@yatsunenko_human_2012]. The additional `r number_string[n-3]` datasets were published after the two reviews were performed and include datasets with more samples than were found in the original studies. Our collection of `r number_string[n]` studies allowed us to largely use the same sequence analysis pipeline for all datasets and relied heavily on the availability of public data and access to metadata that included variables beyond the needs of the original study. To execute this analysis, we created an automated data analysis pipeline, which can be easily updated to add additional studies as they become available. Similarly, it would be possible to adapt this pipeline to other body sites and treatment or variables (e.g. subject's sex or age).

Similar to our study, the Walters et al [@walters_meta-analyses_2014] generated Random Forest machine learning models to differentiate between non-obese and obese individuals. They obtained similar AUC values to our analysis; however, they did not attempt to test these models on the other studies in their analysis. When we performed the inter-dataset cross validation the median accuracy across datasets was only `r 100*median(test_summary$accuracy)`% indicating that the models did a poor job when applied to other datasets. This could be due to differences in subject populations and methods. Considering the median AUC for models trained and tested on the same data with ten-fold cross validation only varied between `r min(genus$auc_cv)` and `r max(genus$auc_cv)` and there was not a strong signal in the alpha diversity data, we suspect that there is insufficient signal to reliably classify individuals.

Although we failed to find an effect this does not necessarily mean that there is no role for the microbiome in obesity. There is strong evidence in murine models of obesity that the microbiome and level of adiposity can be manipulated via genetic manipulation of the animal and manipulation of the community through antibiotics or colonizing germ free mice with diverse fecal material from human donors [@turnbaugh_obesity-associated_2006; @Koren2012; @Cox2014; @Mahana2016; @ley_obesity_2005]. These studies appear to conflict with the observations using human subjects. Recalling the large interpersonal variation in the structure of the microbiome, it is possible that each individual has their own signatures of obesity. Alternatively, it could be that the involvement of the microbiome in obesity is at the level of a common set of metabolites that can be produced from different structures of the microbiome.

@@

@@

**Figure S1: Individual and Combined comparison of Obese and Non-Obese groups Based on Evenness (A), Richness (B), or the Relative Abundance of *Bacteroidetes* (C) and Firmictues (D).**

**Figure S2: Meta Analysis of the Relative Risk of Obesity Based on Evenness (A), Richness (B), or the Relative Abundance of *Bacteroidetes* (C) and Firmictues (D).**

@@

**Figure S6: Power (A) and sample size simulations (B) for the relative abundance of *Bacteroidetes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.

**Figure S7: Power (A) and sample size simulations (B) for the relative abundance of *Firmicutes* for differentiating between non-obese versus obese for effect sizes of 1, 5, 10, and 15%.** Power calculations use the sampling distribution from the original studies and the sample size estimations assume an equal amount of sampling from each treatment group.
